Subsidiary Recursion in Coq

### 2022-01-22 paper/paper.tex
completed revisions based on Chris's comments
Paper is too long now, but I will incorporate Ben's comments and then
shrink.

\begin{figure}
  \begin{verbatim}
Definition List := Subrec ListF .
Definition inList : ListF List -> List := inn ListF.
Definition mkNil : List := inList Nil.
Definition mkCons (hd : A) (tl : List) : List := inList (Cons hd tl).
Definition toList : list A -> List.
Definition fromList : List -> list A.
\end{verbatim}
  \caption{Some basics from \texttt{List.v}, specializing the functor-generic derivation of subsidiary recursion to lists (\texttt{List.v})}
  \label{fig:listf}
\end{figure}

@@

The final definitions in the figure are for \verb|fold|, which allows
us to fold an \verb|Alg| over a value of type \verb|Subrec|; and for mapping
between \verb|Alg| and its unfolding in terms of \verb|AlgF|.  We will
return to the code for \verb|Subrec.v| in Section~\ref{sec:deriv}.

@@

Before we define \verb|spanr|, we must resolve a small problem.
If the first element of the input list
\verb|xs| to \verb|span| does not satisfy \verb|p|, then \verb|span|
should return \verb|([], xs)|.  But when recursing on \verb|xs|, we
will see it only in the form of a subdata structure of type
\verb|F R|.  We will not be able to return it from our recursion at
type \verb|R|, and hence we would not be able to return \verb|([],xs)|
as desired.  To work around this, we will have our recursion return a value
of type \verb|SpanF R| (\verb|X| will be implicit for the constructors):
\begin{verbatim}
Inductive SpanF(X : Set) : Set :=
  SpanNoMatch : SpanF X
| SpanSomeMatch : list A -> X -> SpanF X.
\end{verbatim}
\noindent The idea is that the recursion will signal if it is in the
one tricky case where \verb|p| does not match the first element, by
returning \verb|SpanNoMatch|.  Otherwise, it will be able to return,
via \verb|SpanSomeMatch|, a prefix and the suffix at type
\verb|R|. The prefix will be nonempty, and hence the suffix will be at
most the tail of \verb|xs|.  This tail is available to the algebra
in the subdata structure of type \verb|F R|.

@@

\verb|SpanAlg| is used in the definition of \verb|spanhr|, in
Figure~\ref{fig:span}.  This function invokes the fold function it is
given, on \verb|SpanAlg|.  The final twist is now in the definition of
\verb|spanr|.  We call \verb|spanhr| on the input \verb|xs : R|.  If
\verb|spanhr| returns \verb|SpanNoMatch|, then we are supposed to
return \verb|([],xs)|, which we can do here, because we have
\verb|xs : R|.  It was only inside the algebra that we lost the
information that the subdata structure of type \verb|F R| is derived
from a value of type \verb|R|.  If \verb|spanhr| returns
\verb|SpanSomeMatch l r|, then we return the nonempty
prefix (\verb|l|) and the suffix (\verb|r|).  We also define a version of \verb|break| for subsidiary recursion.

@@

Our work contributes to the program proposed by Owens and Slind, of
broadening the scope of functional programs that can be accommodated
in ITPs~\cite{owens+08}.  The goal of terminating recursion has been
advocated in the literature on programming languages under the name
\emph{strong functional programming}~\cite{turner95}.  Uustalu and
Vene developed a categorical view of a recursion scheme allowing one
level of subsidiary recursion, and illustrated it in Haskell with an
artificial example~\cite{uustalu11}.  In contrast, our scheme allows
arbitrary finite nestings of recursion, and we illustrate it in Coq
with realistic examples.  Like them, we find that subsidiary recursion
subsumes course-of-value recursion.

\textbf{Mendler-style recursion.}  Mendler introduced the idea of using
universal abstraction to support compositional termination checking~\cite{mendler91}.  He proposed a
functor-generic recursor of type $\all{X}{(\all{R}{(R \to X) \to F\ R
    \to X}) \to \mu\ F\ \to X}$. We have adopted this idea to the
constructor of the type \verb|Mu| (Section~\ref{sec:mu}).  Previous
works explored the categorical perspective on Mendler-style
recursion~\cite{uustalu99}, and its use with negative type
schemes~\cite{ahn11}.  Previous work from our group showed
how to derive inductive datatypes in Cedille using encodings
extending the Mendler encoding~\cite{firsov+18b,firsov+18a}.
Here, we do not derive inductive types using these methods,
but rather apply them to justify a terminating recursion scheme
for existing datatypes.

>>>>

\begin{figure}
  \begin{verbatim}
Definition List := Subrec ListF.
Definition inList : ListF List -> List := inn ListF.
Definition mkNil : List := inList Nil.
Definition mkCons (hd : A) (tl : List) : List := inList (Cons hd tl).
Definition toList : list A -> List.
Definition fromList : List -> list A.
\end{verbatim}
  \caption{Some basics from \texttt{List.v}, specializing the functor-generic derivation of subsidiary recursion to lists (\texttt{List.v})}
  \label{fig:listf}
\end{figure}

@@

The last definitions in the figure are for \verb|fold|, which allows
us to fold an \verb|Alg| over a value of type \verb|Subrec|; and for mapping
between \verb|Alg| and its unfolding in terms of \verb|AlgF|.  We will
return to the code for \verb|Subrec.v| in Section~\ref{sec:deriv}.

Finally, for a recursion scheme, one would like to see not just the
typed interface, but also the computation law.  This is stated as a
theorem in Figure~\ref{fig:comp}.  Intuitively, it states that
\verb|fold|ing an algebra over constructed data \verb|inn d| is equal
to invoking the algebra on the identity function for \verb|reveal|
from the interface for algebras (Figure~\ref{fig:algf}); \verb|fold|
for the fold function; an invocation of \verb|fold| with the algebra
for the \verb|eval| function; and \verb|d| for the subdata structure.

\begin{figure}
\begin{verbatim}
Theorem FoldChar :
 forall (X : Set -> Set) (FunX : Functor X) (IdF : FmapId X FunX)
        (algf : AlgF Alg Subrec X) (d : F Subrec),
 fold X FunX (rollAlg algf) (inn d) =
   algf _ (fun x => x) fold (fold X FunX (rollAlg algf)) d .
\end{verbatim}
\end{figure}

@@

Before we define \verb|spanr|, we must resolve a small problem.
If the first element of the input list
\verb|xs| to \verb|span| does not satisfy \verb|p|, then \verb|span|
should return \verb|([], xs)|.  But when recursing on \verb|xs|, we
will see it only in the form of a subdata structure of type
\verb|ListF A R|.  We will not be able to return it from our recursion at
type \verb|R|, and hence we would not be able to return \verb|([],xs)|
as desired.  To work around this, we will have our recursion return a value
of type \verb|SpanF R| (\verb|X| will be implicit for the constructors):
\begin{verbatim}
Inductive SpanF(X : Set) : Set :=
  SpanNoMatch : SpanF X
| SpanSomeMatch : list A -> X -> SpanF X.
\end{verbatim}
\noindent The idea is that the recursion will signal if it is in the
one tricky case where \verb|p| does not match the first element, by
returning \verb|SpanNoMatch|.  Otherwise, it will be able to return,
via \verb|SpanSomeMatch|, a prefix and the suffix at type
\verb|R|. The prefix will be nonempty, and hence the suffix will be at
most the tail of \verb|xs|.  This suffix is available to the algebra
in the subdata structure of type \verb|ListF A R|.

@@

\verb|SpanAlg| is used in the definition of \verb|spanhr|, in
Figure~\ref{fig:span}.  This function invokes the fold function it is
given, on \verb|SpanAlg|.  The final twist is now in the definition of
\verb|spanr|.  We call \verb|spanhr| on the input \verb|xs : R|.  If
\verb|spanhr| returns \verb|SpanNoMatch|, then we are supposed to
return \verb|([],xs)|, which we can do here, because we have
\verb|xs : R|.  It was only inside the algebra that we lost the
information that the subdata structure of type \verb|F R| is derived
from a value of type \verb|R|.  If \verb|spanhr| returns
\verb|SpanSomeMatch l r|, then we return the nonempty
prefix (\verb|l|) and the suffix (\verb|r|).  We also define a version of \verb|break| for subsidiary recursion (e.g., in \verb|wordsBy|, below).

@@

Our work contributes to the program proposed by Owens and Slind, of
broadening the scope of functional programs that can be accommodated
in ITPs~\cite{owens+08}.  The goal of terminating recursion has been
advocated in the literature on programming languages under the name
\emph{strong functional programming}~\cite{turner95}.  Uustalu and
Vene developed a categorical view of a recursion scheme allowing one
level of subsidiary recursion, and illustrated it in Haskell with an
artificial example~\cite{uustalu11}.  In contrast, our scheme allows
arbitrary finite nestings of recursion, and we illustrate it in Coq
with realistic examples.  Like them, we find that subsidiary recursion
subsumes course-of-value recursion.  The technique of sized types is
similar to our method in providing a type-based method for
termination~\cite{barthe04}.  This technique enriches datatypes with
abstract sizes, which must then be propagated through code, using
dependent types.  Our \verb|Subrec| does not require dependent types,
resulting in just polymorphically typed code for the examples we saw
(\verb|Subreci|, for proofs, of course does use dependent types).

\textbf{Mendler-style recursion.}  Mendler introduced the idea of
using universal abstraction to support compositional termination
checking~\cite{mendler91}.  He proposed a functor-generic recursor of
type $\all{X}{(\all{R}{(R \to X) \to F\ R \to X}) \to \mu\ F\ \to
  X}$. We have adopted this idea to the constructor of the type
\verb|Mu| (Section~\ref{sec:mu}).  Previous work explored the
categorical perspective on Mendler-style recursion, and showed how to
reduce it to basic catamorphisms (i.e., structural
recursion)~\cite{uustalu99}.  Another considered its use with negative
type schemes~\cite{ahn11}.  Previous work from our group showed how to
derive inductive datatypes in Cedille using extensions of the Mendler
encoding~\cite{firsov+18b,firsov+18a}.  Here, we do not derive
inductive types using these methods, but rather apply them to justify
a terminating recursion scheme for existing datatypes.
